\chapter{Abstract}

\subsection*{Part~\ref{part:neutron-stars} (project thesis): Neutron Stars}

Neutron stars contain matter at extremely high densities, where both general relativity and
quantum many-body physics are required to describe their structure. In this project, we study
how different models of dense matter affect the relation between the mass and radius of a
neutron star. The analysis is based on the Tolman–Oppenheimer–Volkoff equations, which are solved
numerically for various equations of state.
We first consider a simple model, namely an ideal, non-interacting Fermi gas. This illustrates how
degeneracy pressure can counteract gravitational collapse, while also demonstrating the limitations
of such idealised descriptions when confronted with observed neutron-star properties.
Building on this, the main part of the project develops an interacting model for nucleons and
leptons in equilibrium. The interactions are described through self-consistent meson fields,
leading to a more realistic equation of state, which is combined with a standard crust
model before being used in the numerical stellar calculations.
The resulting mass–radius relations yield radii typical of neutron stars,
but they do not support the heaviest observed stars and show clear differences from current
astronomical constraints. This demonstrates that while the interacting model improves upon the
ideal, non-interacting Fermi gas description, it remains incomplete at the highest densities.
Overall, the project presents a coherent path from ideal models to a description of interacting
dense matter and shows the need for more advanced equations of state in order to reproduce
the full range of observed neutron-star properties.

\subsection*{Part~\ref{part:quark-stars} (Master's thesis): Quark Stars}

This thesis studies compact-star solutions from two-flavor effective quark-matter
models, with the main focus on the Quark--Meson--Diquark model and
two-flavor color superconductivity. The equations of state are constructed in
the mean-field approximation and used as input to the
Tolman--Oppenheimer--Volkoff equations.
The two-flavor Quark--Meson model is first used as a baseline. Varying the sigma
mass changes both the stiffness of the equation of state and the topology of the
stellar branch: lower sigma masses give larger maximum masses and
self-bound-like low-mass behavior, while the \(m_\sigma=600~\mathrm{MeV}\)
case gives a gravitationally bound sequence. A Bodmer--Witten-inspired vacuum
shift reduces the mass and radius scale, but does not qualitatively change this
ordering in the explored range.
The main result is obtained from the Quark--Meson--Diquark model with 2SC
pairing. Retaining the full one-loop potential is essential: removing
\(\Omega_{1,\mathrm{num}}\) can destroy robust 2SC pairing. 
For neutral stellar matter, the baseline
QMD model gives a gravitationally bound sequence with
\(M_{\max}=1.970\,M_\odot\) and \(R(M_{\max})=12.16\,\mathrm{km}\).
Parameter variations show that \(g_\Delta\), \(m_\Delta\), and \(\lambda_3\)
control complementary aspects of the mass--radius relation. In particular,
selected combined parameter choices exceed \(2\,M_\odot\) and pass visually
through the plotted NICER regions for both J0030+0451 and J0740+6620. The
results indicate that two-flavor QMD matter with 2SC pairing can produce
phenomenologically promising compact-star sequences, but the present models are
pure quark-star sequences without a hadronic envelope, strange quarks, tidal
deformabilities, or statistical inference.
