\chapter*{Disclaimer to the External Examiner}

This document contains work from both the project thesis and the master's
thesis, collected in a single volume. It is divided into three main parts:
\begin{enumerate}
    \item Part~\ref{part:neutron-stars} contains the project thesis.
    \item Part~\ref{part:quark-stars} contains the master's thesis.
    \item Part~\ref{part:appendices} contains appendices for both parts.
\end{enumerate}

Only Part~\ref{part:quark-stars} and Appendices~\ref{app:renormalization},
\ref{app:qmd_potential}, and \ref{app:code} are to be evaluated as part of the
master's thesis.

The project thesis is included because it provides important background and
motivation for the master's thesis. In particular, Part~\ref{part:neutron-stars}
develops the compact-star framework used later and shows how the equation of
state affects mass--radius relations. Part~\ref{part:quark-stars} builds on this
foundation by replacing the hadronic description with effective quark-matter
models and studying quark-star solutions.

Keeping both parts in the same document avoids repeating material from the
project thesis and makes it possible to refer directly to earlier definitions,
methods, and results. At the same time, the separation into parts is intended to
make clear which material belongs to the project thesis and which material
belongs to the master's thesis. References from Part~\ref{part:quark-stars} back
to Part~\ref{part:neutron-stars} are thus included for continuity and background.
