\chapter{Introduction}
\label{chap:introduction2}

Neutron stars are compact stellar remnants formed in core-collapse supernovae.
They typically contain masses of order one to two solar masses within radii of
roughly ten kilometres, making them among the densest observable objects in the
universe. Their interiors reach densities far above those of ordinary nuclei,
and the microscopic composition of matter under such conditions remains one of
the central open questions in compact-star physics. Observations from radio
timing, X-ray pulse-profile modeling, and gravitational-wave detections have
therefore turned neutron stars into laboratories for strongly interacting matter
at extreme density \cite{LattimerPrakash2004,glendenning2000,
Riley_2019_J0030,Miller_2019_J0030,Miller_2021_J0740,Riley2021,LIGO2018}.

The structure of a static, spherically symmetric compact star is determined
within general relativity by the Tolman--Oppenheimer--Volkoff (TOV) equations
\cite{tolman1939,oppenheimer1939}. These equations describe hydrostatic
equilibrium by relating the pressure, energy density, and enclosed mass of the
star. To solve them, one must provide an equation of state. The equation of
state encodes the microscopic degrees of freedom and interactions of dense
matter, and it is the main source of theoretical uncertainty in compact-star
modeling. Different assumptions about the equation of state can lead to
substantially different predictions for the maximum mass, radius, central
density, and stability of compact-star configurations.

The importance of the equation of state is already visible in the simplest
models. The classic Oppenheimer--Volkoff calculation, based on an ideal
degenerate neutron gas, gives a maximum mass of approximately
\(0.7\,M_\odot\) \cite{oppenheimer1939}. This is far below the masses of
observed neutron stars and shows that interactions between particles are
essential. More realistic hadronic models include interacting neutrons and
protons, leptons, charge neutrality, and $\beta$-equilibrium. Such models represent
a major improvement over the ideal Fermi-gas description and can support much
larger stellar masses. This progression was studied in Part~\ref{part:neutron-stars} of the thesis,
where compact-star structure was developed from ideal gases to interacting
relativistic mean-field matter. The main lesson from that analysis was that
microphysical assumptions are directly reflected in macroscopic mass--radius
relations.

Even with interacting hadronic matter, however, the equation of state remains
uncertain at the highest densities. Models based on nucleonic degrees of
freedom, such as relativistic mean-field descriptions or microscopic hadronic
equations of state, can reproduce important aspects of neutron-star
phenomenology, but their extrapolation to several times nuclear saturation
density is not unique. A sufficiently stiff equation of state is needed to
support massive compact stars, while radius measurements and gravitational-wave
constraints restrict how stiff the matter can be at lower and intermediate
densities. This tension motivates the study of additional degrees of freedom and
new phases of strongly interacting matter in the cores of compact stars
\cite{Baym2018,Annala2020}.

At sufficiently high density, the relevant microscopic degrees of freedom may no
longer be hadrons, but quarks and gluons. Quantum chromodynamics (QCD) is the
fundamental theory of the strong interaction and describes hadrons in terms of
colored quarks interacting through gluons \cite{GrossWilczek1973,
PeskinSchroeder1995}. Due to asymptotic freedom, QCD becomes weakly coupled at
very large energy or density scales. This suggests that matter at sufficiently
high baryon density may undergo a transition from confined hadronic matter to
deconfined quark matter. Compact stars are among the few physical systems in
which such densities may be realized, making them natural laboratories for
studying possible quark phases of QCD.

The possibility of quark matter in compact stars has a long history. Early work
considered stars containing deconfined quark matter, either as hybrid stars with
quark cores or as stars composed entirely of quark matter
\cite{Itoh1970,Bodmer1971,Witten1984,glendenning2000}. The strange matter
hypothesis, proposed by Bodmer and later developed by Witten, suggests that bulk
three-flavor quark matter could be the true ground state of strongly interacting
matter \cite{Bodmer1971,Witten1984}. If this hypothesis is realized, compact
stars could in principle be strange stars rather than ordinary neutron stars.
In the present thesis, however, the models studied contain only two quark
flavors. The strange matter hypothesis is therefore relevant as motivation, but
the calculations performed here should not be interpreted as a direct test of
absolutely stable strange matter.

A direct first-principles treatment of cold dense QCD remains difficult. Lattice
QCD is highly successful at zero baryon chemical potential, but the sign problem
prevents straightforward simulations of the cold, dense regime relevant for
neutron-star interiors. Perturbative QCD is reliable only at densities much
higher than those expected in the cores of observed compact stars. As a result,
one must often rely on effective models that retain selected symmetries and
degrees of freedom of QCD while remaining tractable. Such models cannot replace
QCD itself, but they provide useful laboratories for studying how chiral symmetry,
deconfinement-inspired quark degrees of freedom, and pairing mechanisms affect
the equation of state and compact-star structure
\cite{Baym2018,Alford2008,Kurkela2010,Komoltsev2022}.

The present thesis studies such effective quark-matter descriptions using the
two-flavor Quark--Meson model and its extension with diquark degrees of freedom.
The Quark--Meson model describes quarks coupled to mesonic fields and provides a
simple framework for studying chiral symmetry breaking and restoration in dense
matter. By constructing the thermodynamic potential and minimizing it in the
mean-field approximation, one obtains an equation of state that can be used as
input to the TOV equations. This makes it possible to study how parameters in
the chiral sector, such as the sigma mass, affect the stiffness of quark matter
and the resulting compact-star mass--radius relations.

The thesis also considers a phenomenological Bodmer--Witten-inspired shift of
the vacuum normalization in the two-flavor Quark--Meson model. This construction
is useful for studying how changes in the zero-pressure point affect quark-star
sequences, in a way similar to a bag-like pressure shift. Since the model is
two-flavor, this should not be understood as a calculation of stable strange
matter. Rather, it is used as a diagnostic tool for exploring how the mass and
radius scales of quark-star solutions respond to changes in the vacuum
normalization.

The main extension studied in this thesis is the two-flavor
Quark--Meson--Diquark (QMD) model. This model includes a diquark condensate and allows
the two-flavor color-superconducting phase to be studied explicitly. Color
superconductivity is expected to be a generic possibility in dense quark matter,
where attractive quark--quark channels can lead to Cooper pairing
\cite{Alford2008}. In the two-flavor case, the relevant paired phase is the 2SC
phase. The QMD model therefore provides a framework for studying how 2SC pairing
modifies the equation of state, the speed of sound, and the compact-star
mass--radius relation.

A central question in this thesis is whether such two-flavor quark-matter models
can produce compact-star sequences compatible with the main qualitative
astrophysical requirements: sufficiently large maximum masses, radii in the
range suggested by X-ray observations, and stable stellar branches. The goal is
not to construct a final realistic neutron-star equation of state. The models
used here do not include a hadronic crust or envelope, a nuclear-to-quark
transition, strange quarks, color-flavor-locked (CFL) pairing, or a Bayesian comparison with
observational data. Instead, the aim is to isolate the physics of the quark
sector and determine which mechanisms and parameters are most important for
producing astrophysically relevant compact-star behavior.

The thesis is organized as follows. Chapter~\ref{chap:qft} introduces the
quantum-field-theoretic tools needed for the effective models used later,
including scalar field theory, interactions, and renormalization. 
Chapter~\ref{chap:qm_model} develops the two-flavor Quark--Meson model, its
mean-field thermodynamics, and the renormalization procedure used to define the
model consistently. Chapter~\ref{chap:quark_matter_stars} applies these models
to dense quark matter and compact stars. It first studies simple Quark--Meson
equations of state and the corresponding quark-star sequences, then introduces
the Bodmer--Witten-inspired shift, and finally analyzes the QMD model with 2SC pairing. The chapter also examines parameter sensitivity and
compares the resulting mass--radius curves with selected astrophysical
constraints.
