\chapter{Two-Flavor Quark--Meson Model}

The modern understanding of strongly interacting matter is rooted in
the development of QCD in the early 1970s.
With the discovery of asymptotic freedom, it became clear that the
fundamental degrees of freedom of the strong interaction are quarks
and gluons, whose effective interactions become weaker at large
momentum scales \cite{GrossWilczek1973,Politzer1973}. This insight
established QCD as a consistent microscopic theory and clarified that
the appropriate degrees of freedom describing strongly interacting
matter need not be the same in all physical regimes. While hadrons
emerge as the relevant low-energy excitations of QCD in vacuum and at
moderate densities, the theory itself is formulated in terms of quarks
and gluons, suggesting that quark-based descriptions may become more
appropriate at sufficiently high baryon densities.

QCD provides the fundamental description of the strong interaction,
but its direct application to cold and dense systems remains highly
challenging. The primary difficulty is that many nonperturbative
methods developed for QCD, including lattice QCD, are well suited to
high temperatures or vanishing baryon chemical potential, but become
problematic at finite density.
In particular, the fermion sign problem obstructs straightforward
numerical simulations in the regime of high baryon density and low
temperature. At finite baryon chemical potential, the mathematical
weight that appears in lattice QCD calculations is no longer positive
and can even become complex. This prevents its interpretation as a
probability distribution and makes standard Monte Carlo simulation
methods impractical \cite{TroyerWiese2005,DeForcrand2010}. As a result,
first-principles lattice calculations of dense QCD matter are
currently not available.

These difficulties have been recognized since the early development
of QCD and have motivated the construction of a variety of effective
models for strongly interacting matter
\cite{CollinsPerry1975,BaymChin1976}. Such models are not intended as
first-principles descriptions. Instead, they are designed to capture
selected physical features of QCD—most notably its symmetries and
symmetry-breaking patterns—while remaining computationally tractable.
In this way, effective models provide a practical framework for
exploring the qualitative behavior of dense matter in regimes where
direct applications of QCD are not yet possible.

The Quark--Meson model is one example of such an effective
description. Historically, it is closely related to the linear sigma
model introduced by Gell-Mann and Lévy as a theory of nucleons
interacting with scalar and pseudoscalar mesons
\cite{GellMannLevy1960}. In later developments, the fermionic degrees
of freedom were reinterpreted as quarks rather than nucleons, leading
to what is commonly referred to as the Quark--Meson model. In this
framework, quarks interact via Yukawa couplings with mesonic fields,
which serve as effective degrees of freedom encoding important
features of the strong interaction.

In its simplest form, the model contains two quark flavors, up and
down, coupled to a scalar field and an isotriplet of pseudoscalar
fields. The mesonic sector is constructed such that the model
reproduces the same global symmetry structure as QCD in the limit of
massless light quarks, where the Lagrangian is invariant under chiral
$SU(2)_L \times SU(2)_R$ transformations and different phases of
matter are distinguished by the realization of this symmetry in the
ground state.

At the same time, the model allows for both spontaneous and explicit
breaking of chiral symmetry. Spontaneous symmetry breaking arises
through the structure of the mesonic potential and is associated with
a nonvanishing vacuum expectation value of the scalar field. Explicit
symmetry breaking is introduced through terms that violate chiral
invariance at the level of the Lagrangian, mimicking the effect of
finite light-quark masses in QCD. As a consequence, a sharp chiral
phase transition present in the chiral limit is replaced by a smooth
crossover when explicit symmetry breaking is included.

The two-flavor Quark--Meson model has been widely employed as a
theoretical laboratory for investigating dense strongly interacting
matter, both on its own and in comparison with related chiral quark
models such as the Nambu--Jona-Lasinio model
\cite{Buballa2005}. Its relatively simple structure makes it
particularly well suited for mean-field treatments and for the
derivation of thermodynamic quantities starting from a Lagrangian
formulation \cite{SchaeferWambach2008}. For these reasons, it serves
as a natural starting point for introducing quark-based equations of
state.

Extensions of the model to include a third quark flavor are also
commonly studied. Three-flavor versions allow strange quarks to
participate and are often used in discussions of strange quark matter
and the possible appearance of strangeness at high density
\cite{SchaeferWambach2008,Buballa2005}. While such extensions introduce
additional parameters and physical uncertainties, they represent a
natural continuation of the two-flavor framework. In the present
chapter, however, we restrict attention to the two-flavor
Quark--Meson model, which already captures the essential conceptual
ingredients needed for the discussions that follow.


\section{From QCD to the Quark--Meson model}

The Quark--Meson model is constructed by isolating those features of
QCD that are expected to be most relevant for
the dynamics of light quarks at low and moderate energies, and by
building an effective field model that reproduces these features in
a simplified setting. In particular, the model is designed to reflect
the global symmetry structure of QCD and its characteristic pattern
of symmetry breaking, while the detailed microscopic dynamics
associated with gluon exchange are not retained explicitly
\cite{WeinbergQFT,PeskinSchroeder1995}.

QCD is a quantum field theory describing quarks interacting through
non-Abelian gauge fields. Its dynamics are defined by the QCD
Lagrangian density
\begin{equation}
\mathcal{L}_{\mathrm{QCD}}
= \sum_{f=1}^{N_f} \bar{q}_f^a\!\left(i\gamma^\mu D_\mu - m_f\right)\!q_f^a
- \frac{1}{4}\,G^A_{\mu\nu} G^{A\,\mu\nu},
\label{eq:LQCD}
\end{equation}
where $q_f^a$ denotes the quark field of flavor $f$ and color
$a=1,2,3$, $m_f$ is the corresponding bare quark mass parameter, and
$G^A_{\mu\nu}$ is the gluon field-strength tensor
\cite{PeskinSchroeder1995,Weinberg1995}.

The quark fields are Dirac spinors carrying both color and flavor
degrees of freedom. For later use, it is convenient to organize the
light-quark sector into a flavor multiplet. Restricting attention to
two flavors, one may write
\begin{equation}
q =
\begin{pmatrix}
u \\ d
\end{pmatrix},
\end{equation}
while keeping the explicit sum over flavors in
Eq.~\eqref{eq:LQCD}.

The interaction between quarks and gluons enters through the
gauge-covariant derivative
\begin{equation}
D_\mu \equiv \partial_\mu - i g\,A_\mu^A T^A,
\label{eq:covder}
\end{equation}
where $g$ is the QCD gauge coupling, $A_\mu^A$ are the gluon fields,
and $T^A$ are the generators of the color gauge group $SU(3)$ in the
fundamental representation. The index $A=1,\ldots,8$ labels the eight
generators (and therefore the eight gluon fields). With the standard
normalization one has
\begin{equation}
\mathrm{tr}(T^A T^B) = \frac{1}{2}\delta^{AB},
\end{equation}
and the generators satisfy the commutation relations
\begin{equation}
[T^A,T^B] = i f^{ABC} T^C,
\end{equation}
where $f^{ABC}$ are the structure constants of $SU(3)$
\cite{PeskinSchroeder1995}.

The final term in Eq.~\eqref{eq:LQCD} is the gluonic kinetic term. It
is written in terms of the non-Abelian field-strength tensor
\begin{equation}
G^A_{\mu\nu}
= \partial_\mu A_\nu^A - \partial_\nu A_\mu^A
+ g f^{ABC} A_\mu^B A_\nu^C .
\label{eq:Fmunu}
\end{equation}
The crucial difference from an Abelian gauge theory is the last term,
which is quadratic in the gauge fields. When inserted into
$G^A_{\mu\nu}G^{A\,\mu\nu}$ it produces three- and four-gluon
self-interactions. These self-interactions are a defining feature of
QCD and are ultimately responsible for its non-perturbative
behavior at low energies \cite{PeskinSchroeder1995,Weinberg1995}.

For our purposes it is useful to separate the ``matter part'' of QCD,
which involves quarks, from the purely gluonic part. Suppressing
explicit color indices, the quark contribution may be written
schematically as
\begin{equation}
\mathcal{L}_{\mathrm{q}}
= \bar{q}\,(i\gamma^\mu D_\mu - M)\,q,
\label{eq:Lqschematic}
\end{equation}
where $q$ denotes the quark field (with flavor components collected
into a multiplet) and $M$ is the mass matrix in flavor space. In the
two-flavor case $M=\mathrm{diag}(m_u,m_d)$. In this chapter we restrict
attention to the two lightest quark flavors, up and down, whose
bare masses are small compared to typical hadronic scales
\cite{PeskinSchroeder1995}.

Equation~\eqref{eq:Lqschematic} makes explicit the elements of QCD that
are relevant for the discussion that follows. In particular, it
exhibits the relativistic fermionic nature of the quark fields and the
structure of the mass term, both of which play a central role in the
symmetry properties of the theory. At this stage, we are not concerned
with the detailed dynamics generated by the gluonic self-interactions
in Eq.~\eqref{eq:Fmunu}, but rather with the global symmetries of the
light-quark sector and how these symmetries depend on the quark mass
matrix.

We begin by considering a global phase transformation of the quark
field,
\begin{equation}
q \rightarrow q' = e^{i\alpha} q ,
\qquad
\bar q \rightarrow \bar q' = \bar q e^{-i\alpha},
\label{eq:U1V}
\end{equation}
with constant $\alpha$. Since the phase factors cancel between $q$ and
$\bar q$, both the kinetic term and the mass term in the quark
Lagrangian are invariant under this transformation,
\begin{equation}
\bar q (i\gamma^\mu D_\mu - M) q
\;\longrightarrow\;
\bar q (i\gamma^\mu D_\mu - M) q .
\end{equation}
This symmetry is referred to as a global $U(1)_V$ symmetry and
corresponds to baryon number conservation. Importantly, $U(1)_V$
remains an exact symmetry of QCD even in the presence of finite quark
masses.

A key concept in this context is chirality. In relativistic quantum
field theory, the Dirac matrix $\gamma^5$ allows one to define the
projection operators
\begin{equation}
P_{L,R} = \frac{1}{2}(1 \mp \gamma^5),
\label{eq:projection_operators}
\end{equation}
where $\gamma^5$ satisfies $(\gamma^5)^2 = 1$ and anticommutes with the
Dirac matrices $\gamma^\mu$. These operators possess several important
properties.

First, they are \emph{projection operators}, meaning that applying
them twice is equivalent to applying them once:
\begin{equation}
P_{L,R}^2
= \frac{1}{4}(1 \mp \gamma^5)(1 \mp \gamma^5)
= \frac{1}{4}(1 \mp 2\gamma^5 + (\gamma^5)^2)
= \frac{1}{2}(1 \mp \gamma^5)
= P_{L,R}.
\end{equation}

Second, the left- and right-handed projectors are \emph{orthogonal},
\begin{equation}
P_L P_R
= \frac{1}{4}(1 - \gamma^5)(1 + \gamma^5)
= \frac{1}{4}(1 - (\gamma^5)^2)
= 0,
\end{equation}
which implies that no spinor component can be simultaneously left-
and right-handed.

Third, the projectors satisfy the \emph{completeness relation},
\begin{equation}
P_L + P_R
= \frac{1}{2}(1 - \gamma^5) + \frac{1}{2}(1 + \gamma^5)
= 1,
\end{equation}
so that any Dirac spinor can be decomposed uniquely into left- and
right-handed parts.

Using the projection operators in \eqref{eq:projection_operators}
the quark field may be decomposed into its left- and right-handed
components,
\begin{equation}
q = q_L + q_R, \qquad
q_{L,R} = P_{L,R} q .
\end{equation}
For massless fermions, the left- and right-handed components correspond
to independent degrees of freedom. This property is specific to
relativistic quantum field theories and has no analogue in
non-relativistic quantum mechanics \cite{Weinberg1995}.

An essential empirical fact is that the vacuum of QCD does not respect
the full chiral symmetry of the massless quark Lagrangian, but instead
realizes this symmetry nontrivially through spontaneous symmetry
breaking, as indicated by the presence of a nonvanishing quark
condensate \cite{Nambu1961,WeinbergQFT}. Any model intended to describe
the light-quark sector must therefore reproduce the global chiral
symmetry of the Lagrangian, while at the same time allowing for its
spontaneous breaking in the vacuum. This observation motivates a
careful examination of the symmetry structure of the quark Lagrangian.

We now turn to the chiral limit, $M \to 0$. In this limit, the quark
Lagrangian simplifies and may be written as
\begin{equation}
\mathcal{L}_{\mathrm{q}}
= \bar{q}_L\, i \gamma^\mu D_\mu\, q_L
+ \bar{q}_R\, i \gamma^\mu D_\mu\, q_R .
\label{eq:Lq_chiral}
\end{equation}
The absence of terms coupling $q_L$ and $q_R$ implies that the left-
and right-handed components may be transformed independently without
changing the form of the Lagrangian. This decoupling is the origin of
the enlarged symmetry structure of the massless theory.

For two light quark flavors, the massless Lagrangian
\eqref{eq:Lq_chiral} is invariant under independent non-Abelian flavor
rotations of the left- and right-handed fields,
\begin{equation}
q_L \rightarrow U_L\, q_L,
\qquad
q_R \rightarrow U_R\, q_R,
\qquad
U_L, U_R \in SU(2)_{L,R},
\end{equation}
which defines the global chiral symmetry group
$SU(2)_L \times SU(2)_R$ \cite{WeinbergQFT,ManoharGeorgi}. Explicitly,
an infinitesimal chiral transformation may be written as
\begin{equation}
q_L \rightarrow \left(1 + i \theta_L^i \frac{\tau^i}{2}\right) q_L,
\qquad
q_R \rightarrow \left(1 + i \theta_R^i \frac{\tau^i}{2}\right) q_R ,
\end{equation}
where $\tau^i$ are the Pauli matrices acting in flavor space and
$\theta_L^i$, $\theta_R^i$ are independent real parameters.

In addition to these non-Abelian flavor rotations, the chiral limit
also admits a global axial phase symmetry. Under an axial
transformation,
\begin{equation}
q \rightarrow q' = e^{i\beta \gamma^5} q ,
\qquad
\bar q \rightarrow \bar q' = \bar q e^{i\beta \gamma^5},
\label{eq:U1A}
\end{equation}
with constant $\beta$, the kinetic term remains invariant because
$\{\gamma^5,\gamma^\mu\}=0$. In terms of left- and right-handed
components, this transformation takes the form
\begin{equation}
q_L \rightarrow e^{-i\beta} q_L,
\qquad
q_R \rightarrow e^{+i\beta} q_R ,
\end{equation}
showing explicitly that $U(1)_A$ acts with opposite phases on $q_L$
and $q_R$.

For nonzero quark masses, the axial symmetry is explicitly broken.
Using
\begin{equation}
\bar q M q = \bar q_L M q_R + \bar q_R M q_L ,
\end{equation}
one finds that the mass term is not invariant under the transformation
\eqref{eq:U1A}. Consequently, $U(1)_A$ is a symmetry only in the chiral
limit $M \to 0$.

At the quantum level, the axial $U(1)_A$ symmetry is further broken by
the axial anomaly, which arises from the non-invariance of the path
integral measure under axial transformations and is related to
topologically nontrivial gluon configurations
\cite{Adler1969,BellJackiw1969}. As a result, $U(1)_A$ is not an exact
symmetry of QCD even in the chiral limit. In the present work, we
neglect the effects of the axial anomaly and focus on the non-Abelian
chiral symmetry, which plays the central role in the construction of
the Quark--Meson model.

Collecting the non-Abelian chiral rotations and the vector and axial
phase symmetries at the classical level, the global symmetry group of
the massless two-flavor quark Lagrangian may be written as
\begin{equation}
U(2)_L \times U(2)_R
\simeq SU(2)_L \times SU(2)_R \times U(1)_V \times U(1)_A .
\end{equation}
For small but nonzero light-quark masses, the chiral
$SU(2)_L \times SU(2)_R$ symmetry and the axial $U(1)_A$ symmetry are
explicitly broken, while the vector $U(1)_V$ symmetry associated with
baryon number conservation remains exact
\cite{WeinbergQFT,DonoghueGolowichHolstein}.

The two-flavor Quark--Meson model provides a minimal and explicit
realization of these requirements. It retains quark degrees of
freedom and implements chiral symmetry directly at the level of the
Lagrangian, while introducing additional mesonic fields that allow
the pattern of chiral symmetry breaking to be represented
explicitly. The strength of both spontaneous and explicit symmetry
breaking is governed by a small set of parameters in the mesonic
sector, making their effects straightforward to identify and vary.
The structure of the model is therefore fixed primarily by symmetry
considerations, which strongly constrain the form of the allowed
interaction terms.

The simplest fermion bilinear that couples left- and right-handed
fields is $\bar{q}_L q_R$ (and its Hermitian conjugate). In order to
construct a chirally invariant interaction, this bilinear must be
multiplied by a bosonic field that transforms appropriately under
$SU(2)_L \times SU(2)_R$. This motivates the introduction of a set of
bosonic fields that carry the same chiral transformation properties
as the quark bilinear $\bar{q}_R q_L$. Historically, this idea traces
back to the linear sigma model of Gell-Mann and L\'evy
\cite{GellMannLevy1960}.

Concretely, one introduces a real scalar field $\sigma$ and a triplet
of real pseudoscalar fields
$\boldsymbol{\pi} = (\pi_1,\pi_2,\pi_3)$. These fields are interpreted
as effective collective degrees of freedom associated with chiral
symmetry breaking and its excitations. They do not correspond to
elementary fields in the QCD Lagrangian, but are effective variables
introduced to represent the dominant low-energy dynamics in a manner
consistent with the underlying symmetries of the theory.

It is convenient to organize the mesonic fields into a matrix-valued
field in flavor space,
\begin{equation}
\Sigma = \sigma + i \boldsymbol{\tau}\cdot\boldsymbol{\pi},
\label{eq:sigma_def}
\end{equation}
where $\boldsymbol{\tau}$ denotes the Pauli matrices. The usefulness of
this construction becomes apparent when considering chiral
transformations. Under a global $SU(2)_L \times SU(2)_R$ symmetry
transformation, the quark fields transform as
\begin{equation}
q_L \rightarrow U_L\, q_L,
\qquad
q_R \rightarrow U_R\, q_R ,
\end{equation}
with $U_L, U_R \in SU(2)_{L,R}$. Requiring the interaction between quarks and
mesons to respect the same symmetry uniquely fixes the transformation
property of the mesonic field to be
\begin{equation}
\Sigma \rightarrow U_L\, \Sigma\, U_R^\dagger .
\end{equation}
With this assignment, the combination
$\bar{q}_L \Sigma q_R + \bar{q}_R \Sigma^\dagger q_L$
is invariant under $SU(2)_L \times SU(2)_R$ transformations. This
provides a natural and symmetry-consistent way of coupling quark
fields to the mesonic degrees of freedom.

Having identified the allowed interaction structure, one may now
construct the Lagrangian density of the two-flavor Quark--Meson model.
It consists of three distinct parts: a fermionic kinetic term, a
chirally invariant Yukawa interaction between quarks and mesons, and
a purely mesonic sector describing the dynamics of the bosonic fields,
\begin{equation}
\mathcal{L}_{\mathrm{QM}} =
\bar{q}\left(i \gamma^\mu \partial_\mu
- g \left[\sigma + i \gamma^5 \boldsymbol{\tau}\cdot\boldsymbol{\pi}
\right]\right) q
+ \mathcal{L}_{\mathrm{mes}} .
\end{equation}
Here $g$ is a Yukawa coupling constant. The appearance of the
$\gamma^5$ matrix reflects the pseudoscalar nature of the pion fields
and ensures the correct transformation properties under parity and
chiral symmetry.

The matrix-valued field $\Sigma$, defined in \eqref{eq:sigma_def},
transforms under $SU(2)_L \times SU(2)_R$ as
\begin{equation}
\Sigma \rightarrow U_L \Sigma U_R^\dagger .
\end{equation}
With this transformation rule, the combination
\begin{equation}
\Sigma^\dagger \Sigma
\end{equation}
is invariant under chiral rotations. As a consequence, any function
constructed from $\Sigma^\dagger \Sigma$ alone automatically respects
the global $SU(2)_L \times SU(2)_R$ symmetry.

The kinetic term of the mesonic fields is therefore uniquely fixed by
symmetry to be
\begin{equation}
\mathcal{L}_{\mathrm{kin}}^{\mathrm{mes}}
= \frac{1}{2} \, \partial_\mu \sigma \, \partial^\mu \sigma
+ \frac{1}{2} \, \partial_\mu \boldsymbol{\pi} \cdot \partial^\mu \boldsymbol{\pi}
= \frac{1}{2} \, \mathrm{Tr}\!\left(
\partial_\mu \Sigma^\dagger \partial^\mu \Sigma
\right),
\end{equation}
since derivatives act locally and do not interfere with the global
chiral transformation properties of $\Sigma$.

The remaining freedom resides in the choice of the mesonic potential.
Requiring locality and renormalizability restricts the potential to a
polynomial of finite order in the fields. In four spacetime
dimensions, this implies that interaction terms of order higher than
four are nonrenormalizable and are therefore omitted at this stage.
The potential must furthermore be constructed from chiral invariants,
which limits the allowed terms to combinations of
$\mathrm{Tr}(\Sigma^\dagger \Sigma)$ and its powers.

At quadratic order, the only invariant is
\begin{equation}
\mathrm{Tr}(\Sigma^\dagger \Sigma)
= \sigma^2 + \boldsymbol{\pi}^2 .
\end{equation}
At quartic order, two distinct invariants may be written,
\begin{equation}
\left[\mathrm{Tr}(\Sigma^\dagger \Sigma)\right]^2,
\qquad
\mathrm{Tr}(\Sigma^\dagger \Sigma \Sigma^\dagger \Sigma) .
\end{equation}
For $SU(2)$, however, these two structures are not independent, but
are proportional to one another as a consequence of trace identities
specific to $2\times2$ matrices. As a result, only a single quartic
interaction term is required in the two-flavor case. (This
simplification does not persist for larger symmetry groups, where
multiple independent quartic invariants may appear.)

Including also a linear term that explicitly breaks chiral symmetry,
the most general renormalizable mesonic potential consistent with
$SU(2)_L \times SU(2)_R$ symmetry may be written as
\begin{equation}
V(\sigma,\boldsymbol{\pi}) =
\frac{m^2}{2}(\sigma^2 + \boldsymbol{\pi}^2)
+ \frac{\lambda}{4}(\sigma^2 + \boldsymbol{\pi}^2)^2
- h\,\sigma .
\end{equation}
The parameters $m^2$ and $\lambda$ determine the shape of the
potential, while the linear term proportional to $h$ introduces
explicit chiral symmetry breaking, mimicking the effect of finite
light-quark masses in QCD. In the limit $h \to 0$, the potential
depends only on the invariant combination $\sigma^2 +
\boldsymbol{\pi}^2$ and is fully symmetric under
$SU(2)_L \times SU(2)_R$ transformations.

Each term in the Quark--Meson Lagrangian thus has a clear origin. The
Yukawa interaction provides a chirally invariant mechanism for
generating effective quark masses, the kinetic term describes
propagating mesonic degrees of freedom, and the potential encodes the
possibility of both spontaneous and explicit symmetry breaking. The
structure of the model is therefore fixed almost entirely by symmetry
and renormalizability, in accordance with the general philosophy of
effective field theory \cite{WeinbergEFT}.
